---
title: "From Dictionaries to LLMs"
subtitle: "Text Analysis in R"
author: "Dariia Mykhailyshyna"
institute: "Kyiv School of Economics  ·  R-Ladies Rome"
date: today
format:
  revealjs:
    theme: [default, kse-theme.scss]
    logo: logo_primary_color.png
    footer: "R-Ladies Rome  ·  Text Analysis in R"
    slide-number: c/t
    transition: fade
    background-transition: fade
    highlight-style: github
    code-line-numbers: false
    code-overflow: wrap
    scrollable: true
    fig-align: center
    fig-width: 9
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
  error: true
---

```{r setup}
#| include: false
library(tidyverse)
library(tidytext)
library(stopwords)
library(wordcloud)
library(wordcloud2)
library(topicmodels)
library(igraph)
library(ggraph)
library(textdata)
library(lubridate)
library(scales)
library(mall)
library(ollamar)
llm_use("ollama", "llama3.2", seed = 42, .silent = TRUE)

kse_primary <- "#003964"
kse_blue    <- "#00BBCE"
kse_green   <- "#A7C539"
kse_red     <- "#F15B43"

theme_kse <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title    = element_text(color = kse_primary, face = "bold"),
      axis.title    = element_text(color = kse_primary),
      panel.grid.minor = element_blank()
    )
}
```

## Outline

1. Why text analysis?
2. The data: EUvsDisinfo claims
3. **Part 1** - the tidytext pipeline
   - Tokens, stopwords, frequencies
   - Dictionary-based sentiment
   - Topic modeling (LDA)
   - Bigrams and word networks
4. **Part 2** - LLMs for text analysis with `mall`
5. Dictionaries vs. LLMs: when to use what?

# Why text analysis? {background-color="#003964"}

## A lot of what we care about is text

- Social media posts, news articles, parliamentary speeches
- Open-ended survey responses, customer reviews
- Court rulings, central bank communications, party manifestos
- Misinformation and propaganda

. . .

We want to **quantify** what is in this text:

- What topics dominate?
- What sentiment is expressed?
- How does this change over time, across speakers, across countries?

## Two big families of methods

::: {.columns}
::: {.column width="50%"}
**Traditional / dictionary-based**

- Tokens, counts, dictionaries
- Transparent, fast, cheap
- Easy to audit
- Struggles with sarcasm, negation, context
:::

::: {.column width="50%"}
**LLM-based**

- General-purpose, flexible
- Handles context and nuance
- Slower, more expensive
- A black box - can hallucinate
:::
:::

. . .

Today: both, on the **same dataset**, so you can see what each gives you.

## What you need to follow along

```{r}
#| eval: false
install.packages(c(
  "tidyverse", "tidytext", "stopwords",
  "wordcloud", "wordcloud2", "topicmodels",
  "igraph", "ggraph", "textdata"
))

# For Part 2 (LLMs)
install.packages(c("mall", "ollamar"))
# And install Ollama: https://ollama.com/download
```

Assumed: working knowledge of R and the tidyverse.

# The data {background-color="#003964"}

## EUvsDisinfo

::: {.columns}
::: {.column width="60%"}
- Project tracking pro-Russian misinformation in EU and Eastern Partnership countries
- Each row = one false claim from an article, plus the rebuttal
- January 2015 - January 2020
- ~7,000 claims
- [kaggle.com/datasets/corrieaar/disinformation-articles](https://www.kaggle.com/datasets/corrieaar/disinformation-articles)
:::

::: {.column width="40%"}
**Why this dataset?**

- Real, messy text
- Strong sentiment signal
- Clear time dimension
- Topically rich (war, EU, sanctions...)
:::
:::

## Loading the data

```{r}
articles <- read_csv("data.csv")

articles |>
  select(claim_published, claim_reviewed) |>
  slice_head(n = 3)
```

# Part 1: the tidytext pipeline {background-color="#003964"}

## Bag of words

The core trick of tidytext:

> One token per row.

Once text is in that shape, all your tidyverse muscle memory works:
`filter`, `group_by`, `count`, `left_join`, `ggplot`.

. . .

A "token" is usually a word, but can also be a sentence, a bigram, an n-gram...

## Tokenization with `unnest_tokens`

```{r}
articles_unnested <- articles |>
  unnest_tokens(word, claim_reviewed, to_lower = TRUE)

nrow(articles_unnested)
articles_unnested |> select(claim_published, word) |> slice_head(n = 5)
```

. . .

One row per word per claim. We went from a few thousand rows to a few hundred thousand.

## Stopwords: the boring words

```{r}
stop_words <- stopwords(language = "en", source = "marimo")
head(stop_words, 20)
```

. . .

The `stopwords` package supports many languages and sources - check the help file before picking one.

## Removing stopwords with `anti_join`

```{r}
stop_words_df <- tibble(word = stop_words)

articles_cleaned <- articles_unnested |>
  anti_join(stop_words_df, by = "word")

nrow(articles_unnested) - nrow(articles_cleaned)
```

. . .

`anti_join` drops every row in `articles_unnested` whose `word` matches one in `stop_words_df`. Clean and fast.

## Custom stopwords

Our data is full of country and region names - we want to look at both with and without them.

```{r}
stop_words_countries <- c(
  stop_words,
  "europe", "russia", "eu", "russian", "united", "states",
  "american", "usa", "syria", "ukraine", "kyiv", "donbass",
  "crimea", "belarus", "poland", "western", "us", "ukrainian",
  "eastern", "west", "donbas", "moscow", "european", "germany",
  "georgia", "ukrainians", "union", "belarusian"
)

stop_words_countries_df <- tibble(word = stop_words_countries)

articles_no_countries <- articles_unnested |>
  anti_join(stop_words_countries_df, by = "word")
```

## Word frequencies

```{r}
top_words <- articles_cleaned |>
  count(word, sort = TRUE)

top_words_no_countries <- articles_no_countries |>
  count(word, sort = TRUE)

top_words |> slice_head(n = 6)
top_words_no_countries |> slice_head(n = 6)
```

## Wordcloud (with country names)

```{r}
#| fig-width: 10
#| fig-height: 6
set.seed(1)
wordcloud(
  words      = top_words$word,
  freq       = top_words$n,
  max.words  = 80,
  random.order = FALSE,
  colors     = c(kse_primary, kse_blue, kse_green, kse_red)
)
```

## Wordcloud (without country names)

```{r}
#| fig-width: 10
#| fig-height: 6
set.seed(1)
wordcloud(
  words      = top_words_no_countries$word,
  freq       = top_words_no_countries$n,
  max.words  = 80,
  random.order = FALSE,
  colors     = c(kse_primary, kse_blue, kse_green, kse_red)
)
```

## `wordcloud2` - interactive version

```{r}
#| fig-width: 10
#| fig-height: 5.5
top_words_no_countries |>
  slice_max(n, n = 150) |>
  wordcloud2(size = 0.9,
             color = rep_len(c(kse_primary, kse_blue, kse_green, kse_red), 150))
```

. . .

Interactive (hover shows frequency); save with `htmlwidgets::saveWidget()`.

## Bar plot - more honest

```{r}
#| fig-width: 10
#| fig-height: 5.5
top_words_no_countries |>
  slice_max(n, n = 15) |>
  ggplot(aes(x = fct_reorder(word, n), y = n)) +
  geom_col(fill = kse_blue) +
  coord_flip() +
  labs(x = NULL, y = "Count",
       title = "Most frequent words (country names removed)") +
  theme_kse()
```

. . .

Wordclouds look great. Bar plots actually let people compare frequencies.

## Sentiment analysis

A way to put a number on the emotional tone of a text.

The simplest approach: **dictionary-based**.

A list of words, each tagged with a sentiment. Join, summarize, plot.

Three lexicons we'll use:

- **AFINN** - integer score from -5 to +5
- **Bing** - binary positive / negative
- **NRC** - 8 emotions + positive / negative

## Joining the dictionaries

```{r}
articles_sent <- articles_no_countries |>
  left_join(get_sentiments("nrc"),   by = "word") |> rename(sent_nrc   = sentiment) |>
  left_join(get_sentiments("bing"),  by = "word") |> rename(sent_bing  = sentiment) |>
  left_join(get_sentiments("afinn"), by = "word") |> rename(sent_afinn = value)

articles_sent |>
  select(word, sent_nrc, sent_bing, sent_afinn) |>
  filter(!is.na(sent_afinn)) |>
  slice_head(n = 5)
```

## NRC: which emotions show up?

```{r}
#| fig-width: 10
#| fig-height: 5.5
articles_sent |>
  filter(!is.na(sent_nrc)) |>
  count(sent_nrc) |>
  ggplot(aes(x = fct_reorder(sent_nrc, n), y = n)) +
  geom_col(fill = kse_primary) +
  coord_flip() +
  labs(x = NULL, y = "Count",
       title = "NRC emotions in the claims") +
  theme_kse()
```

## Bing: same data, different story

```{r}
#| fig-width: 10
#| fig-height: 5
articles_sent |>
  filter(!is.na(sent_bing)) |>
  count(sent_bing) |>
  ggplot(aes(x = sent_bing, y = n, fill = sent_bing)) +
  geom_col() +
  scale_fill_manual(values = c(negative = kse_red, positive = kse_green)) +
  labs(x = NULL, y = "Count", title = "Bing sentiment") +
  theme_kse() + theme(legend.position = "none")
```

. . .

NRC said: mostly positive. Bing says: mostly negative. **The dictionary matters.**

## Which words drive each side?

```{r}
#| fig-width: 11
#| fig-height: 5.5
articles_sent |>
  filter(!is.na(sent_bing)) |>
  count(sent_bing, word) |>
  group_by(sent_bing) |>
  slice_max(n, n = 10) |>
  ungroup() |>
  ggplot(aes(x = reorder_within(word, n, sent_bing), y = n, fill = sent_bing)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~sent_bing, scales = "free") +
  scale_x_reordered() +
  scale_fill_manual(values = c(negative = kse_red, positive = kse_green)) +
  coord_flip() +
  labs(x = NULL, y = NULL) +
  theme_kse()
```

## Sentiment over time (AFINN)

```{r}
#| fig-width: 10
#| fig-height: 5
articles_sent |>
  mutate(month = floor_date(as_date(claim_published), "month")) |>
  group_by(month) |>
  summarise(mean_sent = mean(sent_afinn, na.rm = TRUE)) |>
  ggplot(aes(month, mean_sent)) +
  geom_line(color = kse_blue, linewidth = 0.8) +
  geom_smooth(method = "loess", se = FALSE, color = kse_red) +
  labs(x = NULL, y = "Mean AFINN score",
       title = "How negative did the claims get over time?") +
  theme_kse()
```

## Topic modeling: what is this text *about*?

- Unsupervised: no labels needed
- **LDA** (Latent Dirichlet Allocation) assumes each document is a mixture of topics, each topic a distribution over words
- We pick the number of topics $k$ in advance
- Reference: Tidy Text Mining, ch. 6

## Building a document-term matrix

```{r}
articles_dtm <- articles_cleaned |>
  count(word, `...1`) |>
  cast_dtm(`...1`, word, n)

articles_dtm
```

. . .

Sparse matrix: most words don't appear in most documents.

## Fitting an LDA with `k = 2`

```{r}
#| cache: true
set.seed(16)
lda_model <- LDA(
  articles_dtm,
  k = 2,
  method  = "Gibbs",
  control = list(seed = 16)
)
```

## Pulling out per-topic word probabilities

```{r}
topics <- tidy(lda_model, matrix = "beta")
topics |> slice_head(n = 5)
```

`beta` = probability of word given topic.

## Top words per topic

```{r}
#| fig-width: 11
#| fig-height: 5.5
topics |>
  group_by(topic) |>
  slice_max(beta, n = 10) |>
  ungroup() |>
  mutate(term = reorder_within(term, beta, topic)) |>
  ggplot(aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~topic, scales = "free") +
  scale_y_reordered() +
  scale_fill_manual(values = c(kse_blue, kse_red)) +
  theme_kse()
```

## A sharper view: log-ratio of betas

Many of those top words show up in **both** topics. Log-ratios highlight what's *distinctive*.

```{r}
beta_wide <- topics |>
  mutate(topic = paste0("topic", topic)) |>
  pivot_wider(names_from = topic, values_from = beta) |>
  filter(topic1 > .001 | topic2 > .001) |>
  mutate(log_ratio = log2(topic2 / topic1))
```

## Log-ratio plot

```{r}
#| fig-width: 10
#| fig-height: 5.5
beta_wide |>
  group_by(direction = log_ratio > 0) |>
  slice_max(abs(log_ratio), n = 10) |>
  ungroup() |>
  mutate(term = reorder(term, log_ratio)) |>
  ggplot(aes(log_ratio, term, fill = log_ratio > 0)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c(kse_blue, kse_red)) +
  labs(x = "Log2 ratio (topic 2 / topic 1)", y = NULL) +
  theme_kse()
```

. . .

Topic 1: Europe / EU / West. Topic 2: Syria, war, military.

## Bigrams: pairs of words

So far we've thrown away word order. That hides things like:

- "gross domestic product" vs. "gross"
- "**not** good", "**no** evidence"

. . .

```{r}
articles_bigrams <- articles |>
  unnest_tokens(bigram, claim_reviewed, token = "ngrams", n = 2, to_lower = TRUE)

articles_bigrams |> select(bigram) |> slice_head(n = 5)
```

## Filtering bigram stopwords

No dictionary of stopword *pairs*. Split, filter, rejoin.

```{r}
bigrams_separated <- articles_bigrams |>
  separate(bigram, c("word1", "word2"), sep = " ")

bigrams_filtered <- bigrams_separated |>
  filter(!word1 %in% stop_words_df$word,
         !word2 %in% stop_words_df$word)

bigram_counts <- bigrams_filtered |>
  count(word1, word2, sort = TRUE)

bigram_counts |> slice_head(n = 6)
```

## Bigrams catch sentiment mistakes

Words preceded by "not" usually mean the opposite. Let's quantify how much that biases our AFINN scores.

```{r}
not_words <- bigrams_separated |>
  filter(word1 == "not") |>
  inner_join(get_sentiments("afinn"), by = c(word2 = "word")) |>
  count(word2, value, sort = TRUE) |>
  mutate(contribution = n * value)
```

## "not X" sentiment contributions

```{r}
#| fig-width: 10
#| fig-height: 5.5
not_words |>
  slice_max(abs(contribution), n = 20) |>
  mutate(word2 = reorder(word2, contribution)) |>
  ggplot(aes(contribution, word2, fill = contribution > 0)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c(kse_red, kse_green)) +
  labs(x = "Sentiment × count", y = 'Words after "not"') +
  theme_kse()
```

. . .

Net bias: `r round(sum(not_words$contribution), 1)`. Positive number means we over-estimated positivity.

## Word networks with `ggraph`

```{r}
#| fig-width: 11
#| fig-height: 6
bigram_graph <- bigram_counts |>
  filter(n > 20) |>
  graph_from_data_frame()

set.seed(321)
ggraph(bigram_graph, layout = "fr") +
  geom_edge_link(alpha = 0.4, arrow = arrow(length = unit(2, "mm")),
                 end_cap = circle(2, "mm")) +
  geom_node_point(color = kse_blue, size = 3) +
  geom_node_text(aes(label = name), vjust = 1, hjust = 1, size = 3.5) +
  theme_void()
```

## Part 1 recap

A complete pipeline, all in tidyverse syntax:

1. **Tokenize** with `unnest_tokens`
2. **Clean** with `anti_join` and stopwords
3. **Count** to see what's there
4. **Visualize** with wordclouds and bar plots
5. **Score** sentiment by joining dictionaries
6. **Discover topics** with LDA
7. **Use bigrams** to catch what single words miss

. . .

Limitations: no real understanding of meaning. Sarcasm, negation, context all leak through.

# Part 2: LLMs for text analysis {background-color="#003964"}

## Large language models in 60 seconds

- Models trained to predict the next token in text, on enormous corpora
- They learn patterns: grammar, common facts, style
- They are very good at **flexible, context-dependent** text tasks
- They are also good at **confidently making things up** (hallucination)
- Examples: GPT, Claude, Gemini, Llama, Mistral, Deepseek...

## Cloud vs. local

::: {.columns}
::: {.column width="50%"}
**Cloud (OpenAI, Anthropic, ...)**

- Strongest models
- Pay per token
- Data leaves your machine
- API keys, rate limits
:::

::: {.column width="50%"}
**Local (via Ollama)**

- Smaller open models (Llama, Mistral, Qwen)
- Free, unlimited
- Your data stays put
- Slower, weaker, but often "good enough"
:::
:::

. . .

Today we'll use a **local** model so you can replicate at zero cost.

## Setup: Ollama + `mall`

```{r}
#| eval: false
# 1. Install Ollama once: https://ollama.com/download
# 2. Start the Ollama app/server
# 3. From R:
library(mall)
library(ollamar)

ollamar::pull("llama3.2")   # download the model (~2 GB)
```

. . .

That's the whole infrastructure.

## What `mall` gives you

A tidy interface for LLM-powered text operations. Every function works on a data frame, takes a text column, returns a new column.

| Function | What it does |
|:---|:---|
| `llm_sentiment()` | Classify sentiment of each row |
| `llm_classify()` | Assign one of your custom labels |
| `llm_extract()`  | Pull out structured info (people, products, ...) |
| `llm_summarize()` | Short summary of each row |
| `llm_verify()` | Yes/no question on each row |
| `llm_translate()` | Translate to another language |
| `llm_custom()` | Your own prompt |

## Built-in `reviews` data

```{r}
data("reviews")
reviews
```

. . .

A messy review (#3) is the interesting test case for any sentiment tool.

## Sentiment with `mall`

```{r}
#| cache: true
reviews |>
  llm_sentiment(review)
```

. . .

Notice review #3: the LLM recognizes mixed feelings. A dictionary would just average the scores.

## Custom-label classification

```{r}
#| cache: true
reviews |>
  llm_classify(review, labels = c("positive", "negative", "neutral"))
```

```{r}
#| cache: true
# Or use whatever labels are useful for *your* task:
reviews |>
  llm_classify(review,
               labels = c("product complaint", "shipping complaint",
                          "praise", "question"))
```

. . .

Try writing a dictionary for "shipping complaint". This is where LLMs earn their keep.

## Extracting structured fields

```{r}
#| cache: true
reviews |>
  llm_extract(review, "product")
```

. . .

One line, no regex, no NER model. You can also ask for several at once with `c("product", "brand", "feature")`.

## Summarize text

```{r}
#| cache: true
reviews |> llm_summarize(review, max_words = 5)
```

## Verify a yes/no claim

```{r}
#| cache: true
reviews |> llm_verify(review, "is the customer happy with the purchase")
```

. . .

`llm_verify` returns 1/0 - perfect for creating regression-ready variables from open-ended text.

## Translate

```{r}
#| cache: true
reviews |> llm_translate(review, "italian")
```

## Custom prompts

For anything that doesn't fit a built-in helper:

```{r}
#| cache: true
my_prompt <- paste(
  "Answer a question.",
  "Return only the answer, no explanation.",
  "Acceptable answers are 'yes', 'no'.",
  "Is this customer happy with their purchase?:"
)

reviews |>
  llm_custom(review, my_prompt)
```

. . .

Prompt-engineering tip: be explicit about the **output format** you want. Otherwise you'll spend the afternoon parsing.

## Back to the disinformation data

```{r}
#| cache: true
set.seed(1)
claims_sample <- articles |>
  slice_sample(n = 8) |>
  select(claim_reviewed) |>
  mutate(claim_reviewed = str_trunc(claim_reviewed, 250))

claims_sample |>
  llm_sentiment(claim_reviewed) |>
  llm_classify(claim_reviewed,
               labels = c("NATO", "Ukraine", "EU", "Syria", "other"))
```

. . .

Two structured fields from raw text in two lines. (Sample size kept tiny here - local LLM calls take a few seconds each.)

## A reality check

LLMs are not magic. With a local 3B-parameter model expect:

- ~70-90% accuracy on simple sentiment, against human labels
- Occasional category that doesn't match any of your labels
- Made-up entities ("Ministry of Truth", a country that doesn't exist)
- **Always validate** on a hand-labeled sample

. . .

The right question is not *"does it work?"* but *"does it work better than my dictionary, for less effort than training a real classifier?"*

# Dictionaries vs. LLMs {background-color="#003964"}

## When to reach for which

::: {.columns}
::: {.column width="50%"}
**Dictionaries / tidytext**

- Large corpora, tight budget
- You need full reproducibility
- The task is well-matched to a lexicon
- You need to **defend** every score
- Languages with good resources
:::

::: {.column width="50%"}
**LLMs / `mall`**

- Sarcasm, negation, context matter
- Open-ended, ad-hoc questions
- Small or messy samples
- You need to extract custom fields
- Low-resource languages where dictionaries are thin
:::
:::

## And often: both

A common workflow:

1. **Tidytext** for the cheap, fast overview - what's there?
2. **LLM** for the targeted, hard cases - sarcasm, negation, custom labels, entity extraction
3. **Validate** the LLM with a small dictionary-based or hand-labeled sample

. . .

The two are complements, not substitutes.

# Wrap-up {background-color="#003964"}

## Resources

- [Tidy Text Mining with R](https://www.tidytextmining.com/) - Silge & Robinson
- [`tidytext` documentation](https://juliasilge.github.io/tidytext/)
- [`mall` package](https://mlverse.github.io/mall/) - LLM-powered text ops
- [Ollama](https://ollama.com/) - run local LLMs
- [`ellmer`](https://ellmer.tidyverse.org/) - if you want fuller control over LLM calls
- [Datasets for text mining](https://github.com/EmilHvitfeldt/R-text-data)

## Workshops for Ukraine

::: {.columns}
::: {.column width="60%"}
- Charity R workshop series
- Registration fees support Ukrainian causes
- Past topics: causal inference, geospatial, Bayesian, ML, ...
- Open call for instructors and attendees
- [sites.google.com/view/dariia-mykhailyshyna/main/r-workshops-for-ukraine](https://sites.google.com/view/dariia-mykhailyshyna/main/r-workshops-for-ukraine)
:::

::: {.column width="40%"}
**Grazie, Roma!**

[dmykhailyshyna@kse.org.ua](mailto:dmykhailyshyna@kse.org.ua)

Slides and code: github.com/dariia-m
:::
:::

---

::: {.center}
## Questions? {.center}

**Dariia Mykhailyshyna**

Kyiv School of Economics

[dmykhailyshyna@kse.org.ua](mailto:dmykhailyshyna@kse.org.ua)
:::
